---
title: "Diet Optimization Analysis"
output:
  github_document:
    toc: true
    toc_depth: 2
---



# INTRO

This project directory contains code and data for a Meal Plan Optimization project as part of MSDS 460. 

The premise of this project: Create a linear program that reccomends amounts of foods to eat for n number of consecutive days (7 days, for example). The program should seek to minimize the intake of carboydrates present across all meals. The meals must meet all nutritional requirements as prescribed by The Institue of Medicine (IOM). 

# FORMULATION

Minimize the amount of a given nutrient or set of nutrients, e.g., carbohydrates, present in daily meals subject to upper and lower bounds on various nutritional constraints as prescribed by the IOM. Constraints include upper and lower bounds on daily intake of calories, vitamins, minerals, etc. 

```{r child = 'Formulation.Rmd'}

```


Decision variables:xij= amount (in tons per day) of fuel i used by plant j for the mix.Linear programming model:minimize  z = m1in1jijixc,subject tojm1iijijbx,   j = 1, ..., n,bxen1jm1iijijj,0xij,   i = 1, ..., m,  j = 1, .., n.



# DATA:
- NUTRIENT REQUIREMENT DATA form [wikipedia](https://en.wikipedia.org/wiki/Dietary_Reference_Intake)
- COMMON NUTRIENT COUNTS IN FOODS from [USDA](https://www.ars.usda.gov/northeast-area/beltsville-md-bhnrc/beltsville-human-nutrition-research-center/nutrient-data-laboratory/docs/sr28-download-files/)
    - *Note* the values of the nutriets (in the columns) are per 100g of the food item listed. For example "BUTTER, WITH SALT" has 15.87 g of water per 100 g of Butter with Salt 



# EXAMPLE 1) 7-DAY DIET PLAN. LOW CARB

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(lpSolve)
})
```

## Raw data

```{r}
raw_nutriet_data <- suppressMessages(read_csv("data/Nutrient Data.csv"))
message("Number of food items in raw nutriets data: ",scales::comma(nrow(raw_nutriet_data)))
```

```{r}
message("Number of nutriets in raw nutriets data: ",scales::comma(length(raw_nutriet_data)-2))
```

```{r}
raw_constraints <- suppressMessages(read_csv('data/nutrient-constraints.csv') %>% select(`Original Nutrient Name`:Unit))
message("Number of constraints in raw constraints data: ",scales::comma(nrow(raw_constraints)))
```


## Model Data
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
nutriets <- read_csv("data/nutrient-categories.csv")
constraints <- read_csv("data/nutrient-constraints.csv") %>% select(`Original Nutrient Name`:Unit)
```

```{r}
message("Number of food items in processed food data: ",scales::comma(nrow(nutriets)))
```


## Processing

```{r}
# keep only the cross-section of both lists
keep_names <- inner_join(data_frame(names = names(nutriets))
                         , data_frame(names = constraints$`Nutrient Name`)
                         , by = "names") %>% unlist() %>% as.character()

nutriets <- nutriets %>% select_if(names(.) %in% c("Category",keep_names))
constraints <- constraints %>% filter(`Nutrient Name` %in% keep_names)


# data cleansing -----------------------------------------------------------

# manually adding an upper caloric limit of 2,500 kcals
# 2,000 is RDI for women, 2,500 is for men
constraints <- constraints %>% 
  mutate(`Upper Bound` = ifelse(`Nutrient Name`=="Energ_Kcal", 2500,`Upper Bound`))

# if there is no upper bound, then adding UL as lower bound X2
constraints <- constraints %>% 
  mutate(`Upper Bound` = ifelse(`Upper Bound`<`Lower Bound`, `Lower Bound`*2, `Upper Bound`))


# Remove duplicate nutirents and oddities
nutriets <- nutriets %>% 
  filter(!Category %in% c("EGGS","NUTRITIONAL SUPP FOR PEOPLE W/ DIABETES"
                          , "INF FORMULA. MEAD JOHNSON"
                          ,"SHAKE", "SOY SAU", "GUMS", "SHORTENING CONFECTIONERY"
                          , "TOPPING", "MUSHROOM", "WORTHINGTON STRIPPLES"
                          , "MARGARINE-LIKE SHORTENING", "BREAKFAST BAR", "FST FOODS"
                          , "PAPA JOHN'S 14\" CHS PIZZA", "EGG MIX", "JEW'S EAR"
                          , "INF FOR","BEVERAGE", "SUNFLOWER SD BUTTER","SIDE DISHES"
                          , "CISCO", "INF FORMULA")
  ) %>% 
  filter(!grepl(c("BUTTER|OIL|MARGARINE|FAT|LARD"),Category, ignore.case = TRUE))

# nutrients are all amount per 100 grams of food, so convert to per 1 gram
nutriets <- nutriets %>% mutate_at(vars(`Water_(g)`:`Vit_A_(g)`), funs(./100))
```

## View of model data after processing

### Constraints:
```{r}
knitr::kable(constraints)
```

### Sample of food items (decision variables). First ten nutrients 
```{r}
set.seed(1)
knitr::kable(head(sample_frac(nutriets, 1)[,1:10]))
```


## Run the LP in a loop for n number of days

```{r}

# how many days should we plan? (each day is a loop iteration)
all_days <- 7


foods_used <- c("WATER")
all_results <- list()

for(day in 1:all_days){
  # day <- 1

  # for development purposes, sample the nutrients list. Set to 1 to use full list
  sample_size <- 1
  set.seed(1)
  sample_nutriets <- sample_frac(nutriets, sample_size)
  
  # remove foods that have already been used, other than water
  foods_used <- foods_used[foods_used!="WATER"]
  sample_nutriets <- sample_nutriets %>% filter(!Category %in% foods_used)
  
  # set objective: minimize carbohydrates
  objective_function <- sample_nutriets$`Carbohydrt_(g)`
  
  # initiate LHS constraint matrix
  Left_Hand_Side <- matrix(numeric(nrow(sample_nutriets)), nrow = 1)
  
  # for each constraint, create a row in the LHS matrix
  for(i in 1:nrow(constraints)){
    
    contraint_row <- constraints[i,]
    constraint_name <- contraint_row$`Nutrient Name`
    constraint_lower <- contraint_row$`Lower Bound`
    constraint_upper <- contraint_row$`Upper Bound`
    
    nutirent_column <- sample_nutriets %>% select_at(vars(constraint_name)) %>% unlist() %>% as.numeric() %>% matrix(nrow = 1)
    rownames(nutirent_column) <- constraint_name
    Left_Hand_Side <- rbind(Left_Hand_Side, nutirent_column)
  }
  
  # remove the initialization row at the top
  Left_Hand_Side <- Left_Hand_Side[2:nrow(Left_Hand_Side),]
  
  # direction of the constraint lower and upper bound
  constraint_directions <- c(rep(">=", nrow(Left_Hand_Side))
                             ,rep("<=", nrow(Left_Hand_Side))
  )
  
  # Lower and Upper bounds for RHS
  Right_Hand_Side <- c(constraints$`Lower Bound`
                       , constraints$`Upper Bound`)
  
  # duplicate the LHS matrix since we have both upper and lower bounds
  Left_Hand_Side_all <- rbind(Left_Hand_Side, Left_Hand_Side)
  
  ## check the rows and columns match up:
  # all_equal(nrow(Left_Hand_Side_all)
  #           , length(Right_Hand_Side)
  #           , length(constraint_directions)
  # )
  # 
  # all_equal(length(objective_function)
  #           , ncol(Left_Hand_Side_all)
  # )

  lp_time_start <- Sys.time()
  (LP_Solved <- lp(direction = "min"
                   , objective.in = objective_function
                   , const.mat = Left_Hand_Side_all
                   , const.dir = constraint_directions
                   , const.rhs = Right_Hand_Side
                   , presolve=0
                   , compute.sens=TRUE
                   , all.bin = FALSE
                   #, binary.vec
                   #, all.int=FALSE
                   #, int.vec
  ))
  (lp_time <- Sys.time()-lp_time_start)
  
  
  # record results
  result_objective <- LP_Solved$objval
  
  results <- data_frame(
    Food = sample_nutriets$Category[LP_Solved$solution>0]
    ,`Amount(g)` = LP_Solved$solution[LP_Solved$solution>0]
  )
  
  results$Day <- as.integer(day)
  
  all_results[[day]] <- results
  
  # record foods used so they will be removed from subsequent loop
  foods_used <- c(foods_used, results$Food)
  
  # display info
  message("DAY ",day, ": ", length(results$Food)," items selected. "
          ,scales::percent(sample_size), " of data used. LP completed in "
          ,round(lp_time,2), units(lp_time))
}

```


## View results

All amounts are in grams per day. 

For reference:
- 1 cup = 320 grams
- 1 liter = 1000 grams (for water)

```{r paged.print=FALSE}
# print results:
max_len <- max(map_dbl(all_results, nrow))
blank_rows <- data_frame(rownum = 1:max_len)

all_results_print <- all_results %>% 
  map(~.x %>% mutate(rownum = row_number())) %>% 
  map(~full_join(.x, blank_rows, by = "rownum")) %>% 
  map(~.x %>% select(-rownum)) %>% 
  map(~{
    nms <- names(.x)
    dy <- unique(.x$Day)
    dy <- dy[!is.na(dy)]
    nm1 <- paste0("Day ",dy," food")
    nm2 <- paste0("Day ",dy," amt(g)")
    names(.x) <- c(nm1, nm2, "Day")
    .x %>% select(-Day)
  }) %>% 
  bind_cols()

knitr::kable(all_results_print[,1:6])
```


```{r message=FALSE, warning=FALSE, paged.print=FALSE}
knitr::kable(all_results_print[,7:14])
```

# EXAMPLE 2) 1-DAY, LOW SODIUM, LOW CHOLESTEROL, LOW CARBS
Includes a sensitivity analysis

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

# LOW SODIUM, CARBS AND CHOLESTEROL

library(tidyverse)
library(lpSolve)

nutriets <- read_csv("data/nutrient-categories.csv")
constraints <- read_csv("data/nutrient-constraints.csv") %>% select(`Original Nutrient Name`:Unit)

# keep only the cross-section of both lists
keep_names <- inner_join(data_frame(names = names(nutriets))
                         , data_frame(names = constraints$`Nutrient Name`)
                         , by = "names") %>% unlist() %>% as.character()

nutriets <- nutriets %>% select_if(names(.) %in% c("Category",keep_names))
constraints <- constraints %>% filter(`Nutrient Name` %in% keep_names)


# data cleansing -----------------------------------------------------------

# manually adding an upper caloric limit of 2,500 kcals
# 2,000 is RDI for women, 2,500 is for men
constraints <- constraints %>% 
  mutate(`Upper Bound` = ifelse(`Nutrient Name`=="Energ_Kcal", 2500,`Upper Bound`))

# if there is no upper bound, then adding UL as lower bound X2
constraints <- constraints %>% 
  mutate(`Upper Bound` = ifelse(`Upper Bound`<`Lower Bound`, `Lower Bound`*2, `Upper Bound`))


# Remove duplicate nutirents and oddities
nutriets <- nutriets %>% 
  filter(!Category %in% c("EGGS","NUTRITIONAL SUPP FOR PEOPLE W/ DIABETES"
                          , "INF FORMULA. MEAD JOHNSON"
                          ,"SHAKE", "SOY SAU", "GUMS", "SHORTENING CONFECTIONERY"
                          , "TOPPING", "MUSHROOM", "WORTHINGTON STRIPPLES"
                          , "MARGARINE-LIKE SHORTENING", "BREAKFAST BAR", "FST FOODS"
                          , "PAPA JOHN'S 14\" CHS PIZZA", "EGG MIX", "JEW'S EAR"
                          , "INF FOR","BEVERAGE", "SUNFLOWER SD BUTTER","SIDE DISHES"
                          , "CISCO", "INF FORMULA","BEVERAG","TODDL FORM","SHORTENING FRYING (HVY DUTY)"
                          , "SHORTENING INDUSTRIAL")
  ) %>% 
  filter(!grepl(c("BUTTER|OIL|MARGARINE|FAT|LARD"),Category, ignore.case = TRUE))

# nutrients are in grams of nutrient per 100g of food, so reducing to nutrients per 1 gram of food
nutriets <- nutriets %>% mutate_at(vars(`Water_(g)`:`Vit_A_(g)`), funs(./100))

# Run the LP --------------------------------------------------------------

# how many days to plan? (will loop over)
all_days <- 1


foods_used <- c("WATER") # list of foods already used (water is a placeholder)
all_results <- list() # store the meal plans
keep_LPs <- list() # store the LP results
for(day in 1:all_days){
  # day <- 1
  
  # for development purposes, sample the nutrients list. 
  sample_size <- 1
  set.seed(1)
  sample_nutriets <- sample_frac(nutriets, sample_size)
  
  # remove foods that have already been used other than water
  foods_used <- foods_used[foods_used!="WATER"]
  sample_nutriets <- sample_nutriets %>% filter(!Category %in% foods_used)
  
  # set objective: minimize carbohydrates
  objective_function <- sample_nutriets$`Sodium_(g)`+sample_nutriets$`Carbohydrt_(g)`+sample_nutriets$`Cholestrl_(g)` #OBJECTIVE FUNCTION
  
  # initiate LHS matrix
  Left_Hand_Side <- matrix(numeric(nrow(sample_nutriets)), nrow = 1)
  
  # for each constraint, create a row in the LHS matrix
  for(i in 1:nrow(constraints)){
    # i <- 1
    contraint_row <- constraints[i,]
    constraint_name <- contraint_row$`Nutrient Name`
    constraint_lower <- contraint_row$`Lower Bound`
    constraint_upper <- contraint_row$`Upper Bound`
    
    nutirent_column <- sample_nutriets %>% select_at(vars(constraint_name)) %>% unlist() %>% as.numeric() %>% matrix(nrow = 1)
    rownames(nutirent_column) <- constraint_name
    Left_Hand_Side <- rbind(Left_Hand_Side, nutirent_column)
  }
  
  # remove the initialization row at the top
  Left_Hand_Side <- Left_Hand_Side[2:nrow(Left_Hand_Side),]
  
  # there are both lower and upper bounds on the constraints
  constraint_directions <- c(rep(">=", nrow(Left_Hand_Side))
                             ,rep("<=", nrow(Left_Hand_Side))
  )
  
  # Lowe and Upper bounds for RHS
  Right_Hand_Side <- c(constraints$`Lower Bound`
                       , constraints$`Upper Bound`)
  
  # duplicate the LHS matrix since we have both upper and lower bounds
  Left_Hand_Side_Lower <- Left_Hand_Side
  rownames(Left_Hand_Side_Lower) <- paste0(rownames(Left_Hand_Side),"_Lower")
  Left_Hand_Side_Upper <- Left_Hand_Side
  rownames(Left_Hand_Side_Upper) <- paste0(rownames(Left_Hand_Side),"_Upper")
  
  Left_Hand_Side_all <- rbind(Left_Hand_Side_Lower, Left_Hand_Side_Upper)
  
  # check the rows and columns match up:
  all_equal(nrow(Left_Hand_Side_all)
            , length(Right_Hand_Side)
            , length(constraint_directions)
  )
  
  all_equal(length(objective_function)
            , ncol(Left_Hand_Side_all)
  )
  
  
  lp_time_start <- Sys.time()
  (LP_Solved <- lp(direction = "min"
                   , objective.in = objective_function
                   , const.mat = Left_Hand_Side_all
                   , const.dir = constraint_directions
                   , const.rhs = Right_Hand_Side
                   , presolve=0
                   , compute.sens=TRUE
                   , all.bin = FALSE
                   #, binary.vec
                   #, all.int=FALSE
                   #, int.vec
  ))
  (lp_time <- Sys.time()-lp_time_start)
  
  # compute results
  results <- data_frame(
    Food = sample_nutriets$Category[LP_Solved$solution>0]
    ,`Amount(g)` = LP_Solved$solution[LP_Solved$solution>0]
  )
  results$Day <- as.integer(day)
  
  # store results
  keep_LPs[[day]] <- LP_Solved
  all_results[[day]] <- results
  foods_used <- c(foods_used, results$Food)
  
  message("DAY ",day, ": ", length(results$Food)," items selected. ",scales::percent(sample_size), " of data used. LP completed in ",round(lp_time,2), units(lp_time))
}


# print results:
max_len <- max(map_dbl(all_results, nrow))
blank_rows <- data_frame(rownum = 1:max_len)

all_results_print <- all_results %>% 
  map(~.x %>% mutate(rownum = row_number())) %>% 
  map(~full_join(.x, blank_rows, by = "rownum")) %>% 
  map(~.x %>% select(-rownum)) %>% 
  map(~{
    nms <- names(.x)
    dy <- unique(.x$Day)
    dy <- dy[!is.na(dy)]
    nm1 <- paste0("Day ",dy," food")
    nm2 <- paste0("Day ",dy," amt(g)")
    names(.x) <- c(nm1, nm2, "Day")
    .x %>% select(-Day)
  }) %>% 
  bind_cols()

# sensitivity analysis ----------------------------------------------------

# Question: In the event the constraints (i.e., the guidlines issued by IOM) are changed, how much 
# can the RDI's change without alterting the meal plans?

# objective functions 
lp_analysis <- keep_LPs[[1]]

# The dual price gives the improvement in the objective function if the constraint is relaxed by one unit
# "Dual price", also known as Shadow Price

# The shadow price for a constraint indicates the amount by which the 
# objective function value changes given a unit increase in the RHS 
# value of the constraint, assuming all other coefficients remain constant. 
# If a shadow price is positive, a unit increase in the RHS value of the 
# associated constraint results in an increase in the optimal objective function value.

## Sensitivity Report
# 1) Objective
Objective_df <- data_frame(Name = "Minimization of Carbs, Sodium and Cholesterol"
                           , final_value = lp_analysis$objval)

# 2) variables

variable_duals <- data_frame(variable = sample_nutriets$Category
                             , final_value = lp_analysis$solution
                             , objective_coeficiet = objective_function
                             , allowable_increase = lp_analysis$sens.coef.to[(lp_analysis$const.count+1):length(lp_analysis$duals)]
                             , allowable_decrease = lp_analysis$sens.coef.from[(lp_analysis$const.count+1):length(lp_analysis$duals)]
) %>% 
  mutate(Reduced_Cost = objective_coeficiet - lp_analysis$duals[(lp_analysis$const.count+1):length(lp_analysis$duals)]) %>% 
  select(variable, final_value, Reduced_Cost, everything())

variable_duals %>% filter(final_value>0)


# 3) constraints
final_nutrients <- 
  sample_nutriets %>% 
  mutate(solution = lp_analysis$solution) %>% 
  mutate_at(vars(`Water_(g)`:`Vit_A_(g)`), funs(.*solution)) %>% 
  filter(solution>0) %>% 
  select(-solution, -Category) %>% 
  summarise_all(sum) %>% 
  gather(constraint, final_value)

all_final_values <- bind_rows(
  final_nutrients %>% mutate(constraint = paste0(constraint,"_Lower"))
  , final_nutrients %>% mutate(constraint = paste0(constraint,"_Upper"))
  )

Constraint_Sensitivity <- data_frame(constraint = colnames(lp_analysis$constraints)
                                     , shadow_price = lp_analysis$duals[1:lp_analysis$const.count]
                                     , constraint_RHS = Right_Hand_Side
                                     , allowable_increase = lp_analysis$duals.to[1:lp_analysis$const.count]
                                     , allowable_decrease = lp_analysis$duals.from[1:lp_analysis$const.count]
                                     ) %>% 
  left_join(all_final_values, by = "constraint") %>% 
  select(constraint, final_value, everything())

```

### Objective Cell (Min)
```{r}
knitr::kable(Objective_df)
```

### Decision Variable Cells
```{r}
knitr::kable(variable_duals %>% filter(final_value>0))
```


### Constraints
```{r}
knitr::kable(Constraint_Sensitivity %>% filter(shadow_price>0))
```

